---
title: "01_preprocessing"
author: "Martin Proks"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='../reports/')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
  html_document:
    number_sections: yes
    theme: yeti
    toc: yes
    toc_float: yes
    df_print: paged
    dev: png
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = TRUE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("../reports/figures/", DOCNAME, "/"),
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

# Introduction

Basic analysis of scRNA-seq experiment.

- filtering was adjusted
- HVG: 2,500
- regressing cell cycle
- resolution is set to default
- markers include also gene_id and description

```{r, message=FALSE}
library(glue)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
source("../scripts/helpers.R")

# multithreading
library(future)
plan()

options(future.globals.maxSize = 1000 * 1024^2)
plan("multiprocess", workers = 10)
plan()
```

# Load dataset

```{r message=FALSE}
rna <- Read10X('../data/processed/200910_A00642_0101_AHTJW3DRXX-5000_cells/outs/filtered_feature_bc_matrix/', gene.column=2)
colnames(rna) <- stringr::str_replace(colnames(rna), '-[0-9]', '')

glue("Duplicated cell tags: {sum(duplicated(colnames(rna)))}")
raw_ann <- CreateSeuratObject(rna)
```

# Quality control

```{r}
raw_ann[['percent.mito']] <- PercentageFeatureSet(raw_ann, pattern = "^MT-")
raw_ann[['percent.ribo']] <- PercentageFeatureSet(raw_ann, pattern = "^RP[LS]")
```

```{r}
VlnPlot(raw_ann, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mito", "percent.ribo"),
        ncol = 4)
```

```{r}
ggplot(raw_ann@meta.data, aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mito)) + 
  geom_point() + 
  scale_color_continuous(type = "viridis")

nbins <- 100
min_cells <- 6000
max_cells <- 120e3
min_genes <- 1000
max_genes <- 11500
max_mito <- 30

ggplot(raw_ann@meta.data, aes(x=nCount_RNA)) + 
  geom_histogram(bins = nbins) + 
  geom_vline(aes(xintercept=min_cells), linetype="dashed", color='red')

ggplot(raw_ann@meta.data, aes(x=nFeature_RNA)) + 
  geom_histogram(bins = nbins) +
  geom_vline(aes(xintercept=min_genes), linetype="dashed", color='red')

ggplot(raw_ann@meta.data, aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mito)) + 
  geom_point() + 
  scale_color_continuous(type = "viridis") +
  geom_vline(aes(xintercept=min_cells), linetype="dashed", color='red') +
  geom_vline(aes(xintercept=max_cells), linetype="dashed", color='red') +
  geom_hline(aes(yintercept=min_genes), linetype="dashed", color='red') +
  geom_hline(aes(yintercept=max_genes), linetype="dashed", color='red')

ggplot(raw_ann@meta.data, aes(x=percent.mito)) + 
  geom_histogram(bins = nbins) + 
  geom_vline(aes(xintercept=max_mito), linetype="dashed", color='red')
```

## Filtering

```{r}
print(paste0("Before filtering: ", dim(raw_ann)[2], " cells ",  dim(raw_ann)[1], " genes"))
```

```{r}
adata <- subset(raw_ann, subset = 
                    nFeature_RNA > min_genes & nFeature_RNA < max_genes & 
                    nCount_RNA > min_cells & nCount_RNA < max_cells &
                    percent.mito < max_mito)
adata@misc$annotation <- get_annotation(raw_ann)
```

```{r}
ggplot(adata@meta.data, aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mito)) + 
  geom_point() + scale_color_continuous(type = "viridis")

VlnPlot(adata, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mito", "percent.ribo"),
        ncol = 4)
```

```{r}
adata <- CreateSeuratObject(adata@assays$RNA@counts, min.cells = 3, meta.data = adata@meta.data)
```

```{r}
print(paste0("After filtering: ", dim(adata)[2], " cells ",  dim(adata)[1], " genes"))
```

# Normalization

```{r, fig.height=8}
adata <- NormalizeData(adata)
adata <- FindVariableFeatures(adata, nfeatures = 2500)

top10 <- head(VariableFeatures(adata), 10)
plot1 <- VariableFeaturePlot(adata)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

adata <- ScaleData(adata, features = rownames(adata))
```

# Cell Cycle (CC)

## Visualize CC on PCA

Plot below shows that the cell cycle separates specifically due to its phase. This means we have to regress out CC.

```{r}
adata <- CellCycleScoring(adata, s.features = cc.genes$s.genes, g2m.features = cc.genes$g2m.genes, set.ident = TRUE)
adata <- RunPCA(adata, features = c(cc.genes$s.genes, cc.genes$g2m.genes))
DimPlot(adata)
```

## Regress CC

```{r}
adata <- ScaleData(adata, vars.to.regress = c("S.Score", "G2M.Score"), features = rownames(adata))
```

```{r}
adata <- RunPCA(adata, features = c(cc.genes$s.genes, cc.genes$g2m.genes))
DimPlot(adata)
```

# PCA

```{r}
adata <- RunPCA(adata, features = VariableFeatures(object = adata), seed.use = random_seed)
DimPlot(adata, reduction = "pca")
ElbowPlot(adata, ndims = 50)
```

# Clustering

```{r}
adata <- FindNeighbors(adata, dims = 1:30)
adata <- FindClusters(adata, random.seed = random_seed)
```

# Visualization

```{r}
adata <- RunUMAP(adata, dims = 1:20, seed.use = random_seed)
```

## UMAP

```{r}
DimPlot(adata, reduction = "umap", group.by = 'seurat_clusters')
DimPlot(adata, reduction = "umap", group.by = 'Phase')
```

```{r}
DimPlot(adata, reduction = "umap")
FeaturePlot(adata, features = 'NEUROG3', reduction = 'umap')
FeaturePlot(adata, features = 'PDX1', reduction = 'umap')
```

# Markers

```{r}
markers <- FindAllMarkers(adata, logfc.threshold = 1, only.pos = TRUE)
m <- markers %>%
 group_by(cluster) %>%
 top_n(n = 50, wt = avg_log2FC)
m <- cbind(m, adata@misc$annotation[m$gene, c("ensembl_id", "description")])
write.csv(m, file = '../results/01_markers.csv')

m
```

## Save RDS

```{r}
saveRDS(adata, file = '../data/processed/01_adata.RDS')
```
