---
title: "03_subclustering"
author: "Martin Proks"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='../reports/')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
  html_document:
    number_sections: yes
    theme: yeti
    toc: yes
    toc_float: yes
    df_print: paged
    dev: png
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = TRUE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("../reports/figures/", DOCNAME, "/"),
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

# Introduction

Subcluster the left part of UMAP and identify markers.

```{r}
library(dplyr)
library(glue)
library(Seurat)
library(patchwork)
library(ggplot2)
source("../scripts/helpers.R")

# multithreading
library(future)
plan()

options(future.globals.maxSize = 1000 * 1024^2)
plan("multiprocess", workers = 10)
plan()

options(bitmapType='cairo')
```

# Subset

```{r}
adata <- readRDS("../data/processed/02_rna_hto.RDS")
DefaultAssay(adata) <- "RNA"
Idents(adata) <- "seurat_clusters"
```

```{r}
# selected_cells <- CellSelector(DimPlot(adata))
# write.csv(selected_cells, file = "../results/03_subset_cells.csv")
selected_cells <- read.csv("../results/03_subset_cells.csv")$x
adata_sub = adata[, selected_cells]

DimPlot(adata_sub)
```

# Subcluster

```{r}
adata_sub <- FindNeighbors(adata_sub, dims = 1:30)
adata_sub <- FindClusters(adata_sub, resolution = 1, random.seed = random_seed)
```

```{r}
DimPlot(adata_sub)
```

```{r}
saveRDS(adata_sub, file = '../data/processed/03_subcluster.RDS')
```

```{r}
library(SeuratDisk)

adata_sub@assays$RNA@scale.data <- matrix()
SaveH5Seurat(adata_sub, filename = "../data/processed/03_subcluster.h5Seurat", overwrite = T)
Convert("../data/processed/03_subcluster.h5Seurat", dest = "h5ad", overwrite = T)
```

# Markers

## RNA

```{r}
DefaultAssay(adata_sub) <- "RNA"
rna_markers <- FindAllMarkers(adata_sub, logfc.threshold = 1, only.pos = TRUE)
rna_markers <- cbind(rna_markers, adata_sub@misc$annotation[rna_markers$gene, c("ensembl_id", "description")])
write.csv(rna_markers, file = '../results/03_rna_markers.csv')
```

## HTO

```{r}
DefaultAssay(adata_sub) <- "HTO"
hto_markers <- FindAllMarkers(adata_sub, only.pos = TRUE)
write.csv(hto_markers, file = '../results/03_hto_markers.csv')
```

## HTO - based on specific hashtag combinations

- A0262 vs. A0263
- A2062 vs. A0260
- A0263 vs. A0260
- A0260 vs. A0259

```{r}
DefaultAssay(adata_sub) <- "RNA"
Idents(adata_sub) <- "HTO_classification"

m_262_263 <- FindMarkers(adata_sub, ident.1 = "A0262", ident.2 = "A0263", only.pos = F) %>% 
  filter(abs(avg_log2FC) > 0.5 & p_val_adj < 0.05)
m_262_263 <- cbind(m_262_263, adata_sub@misc$annotation[rownames(m_262_263), c("ensembl_id", "description")])

m_262_260 <- FindMarkers(adata_sub, ident.1 = "A0262", ident.2 = "A0260", only.pos = F) %>%
  filter(abs(avg_log2FC) > 0.5 & p_val_adj < 0.05)
m_262_260 <- cbind(m_262_260, adata_sub@misc$annotation[rownames(m_262_260), c("ensembl_id", "description")])

m_263_260 <- FindMarkers(adata_sub, ident.1 = "A0263", ident.2 = "A0260", only.pos = F) %>%
  filter(abs(avg_log2FC) > 0.5 & p_val_adj < 0.05)
m_263_260 <- cbind(m_263_260, adata_sub@misc$annotation[rownames(m_263_260), c("ensembl_id", "description")])

m_260_259 <- FindMarkers(adata_sub, ident.1 = "A0260", ident.2 = "A0259", only.pos = F) %>% 
  filter(abs(avg_log2FC) > 0.5 & p_val_adj < 0.05)
m_260_259 <- cbind(m_260_259, adata_sub@misc$annotation[rownames(m_260_259), c("ensembl_id", "description")])

write.csv(m_262_263, file = "../results/03_A0262_A0263.csv")
write.csv(m_262_260, file = "../results/03_A0262_A0260.csv")
write.csv(m_263_260, file = "../results/03_A0263_A0260.csv")
write.csv(m_260_259, file = "../results/03_A0260_A0259.csv")
```

## Per louvain and HTO

We identify markers by grouping cells based on predefined clustering using louvain.
For each cluster we ask what are the markers in that cluster based on HTO.
Markers generated below are only specific for particular louvain cluster per HTO hashtag.

```{r}
markers <- NULL
for (cluster in unique(adata_sub$seurat_clusters)) {
  for (hashtag in unique(adata_sub$HTO_classification)) {
    m <- NULL
    tryCatch({
      m <- FindMarkers(adata_sub[, adata_sub$seurat_clusters == cluster], 
                       ident.1 = hashtag, group.by = 'HTO_classification', verbose = F)
      m <- m %>% filter(p_val_adj < 0.05 & abs(avg_log2FC) > 1)
      m$cluster <- cluster
      m$hashtag <- hashtag
      m$combined <- glue('{cluster}_{hashtag}')
      markers <- rbind(markers, m)
    }, error = function(error_condition) {
        print(glue('Skipping {hashtag} for cluster #{cluster}'))
    })
  }
}
markers <- cbind(markers, adata_sub@misc$annotation[rownames(markers), c("ensembl_id", "description")])
write.csv(markers, '../results/03_rna_per_hto_markers.csv')
```

# Session info

```{r session-info, cache = FALSE}
devtools::session_info()
```
